---
title: "hpgltools examples using pasilla"
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
 html_document:
  code_download: true
  code_folding: show
  fig_caption: true
  fig_height: 7
  fig_width: 7
  highlight: default
  keep_md: false
  mode: selfcontained
  number_sections: true
  self_contained: true
  theme: readable
  toc: true
  toc_float:
    collapsed: false
    smooth_scroll: false
vignette: >
  %\VignetteIndexEntry{d-04_pasilla}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
  body .main-container {
    max-width: 1600px;
  }
</style>

```{r options, include = FALSE}
library("hpgltools")
knitr::opts_knit$set(progress = TRUE,
                     verbose = TRUE,
                     width = 90,
                     echo = TRUE)
knitr::opts_chunk$set(error = TRUE,
                      fig.width = 8,
                      fig.height = 8,
                      dpi = 96)
old_options <- options(digits = 4,
                       stringsAsFactors = FALSE,
                       knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
set.seed(1)
ver <- "20170820"
rmd_file <- "d-04_pasilla.Rmd"
```

# Pasilla vignette: `r ver`

# Example hpgltool usage with a real data set (pasilla)

In this document, I am hoping to mostly copy/paste material from the tests/ tree and explain the
various functionalities therein.  It is my hope therefore to step from data loading all the way
through ontology searching with appropriate visualizations at each stage.

# Load Data

In test_01load_data.R I perform load some data into an expressionset and get ready to play with it.

```{r load_data}
## I use sm to keep functions from printing too much (well, anything really)
tt <- sm(library(hpgltools))
tt <- sm(library(pasilla))
tt <- sm(data(pasillaGenes))
```

## Gather annotation data

biomart is an excellent resource for annotation data, but it is entirely too complex.
The following function 'get_biomart_annotations()' attempts to make that relatively simple.

```{r biomart}
## Try loading some annotation information for this species.
gene_info_lst <- sm(load_biomart_annotations(species = "dmelanogaster",
                                             host = "useast.ensembl.org"))
gene_info <- gene_info_lst[["annotation"]]
info_idx <- gene_info[["gene_biotype"]] == "protein_coding"
gene_info <- gene_info[info_idx, ]
rownames(gene_info) <- make.names(gene_info[["ensembl_gene_id"]], unique = TRUE)
head(gene_info)
```

## Load count tables

The pasilla data set provides count tables in a tab separated file, let us read them into an
expressionset in the following block along with creating an experimental design.  create_expt() will
then merge the annotations, experimental design, and count tables into an expressionset.

```{r load_counts}
## This section is copy/pasted to all of these tests, that is dumb.
datafile <- system.file("extdata/pasilla_gene_counts.tsv", package = "pasilla")
## Load the counts and drop super-low counts genes
counts <- read.table(datafile, header = TRUE, row.names = 1)
counts <- counts[rowSums(counts) > ncol(counts),]
## Set up a quick design to be used by cbcbSEQ and hpgltools
design <- data.frame(row.names = colnames(counts),
    condition = c("untreated","untreated","untreated",
        "untreated","treated","treated","treated"),
    libType = c("single_end","single_end","paired_end",
        "paired_end","single_end","paired_end","paired_end"))
metadata <- design
colnames(metadata) <- c("condition", "batch")
metadata[["sampleid"]] <- rownames(metadata)

## Make sure it is still possible to create an expt
pasilla_expt <- sm(create_expt(count_dataframe = counts, metadata = metadata,
                               savefile = "pasilla", gene_info = gene_info))
```

# Graph metrics

In this block I will use a single function graph_metrics() to plot them all.
And then follow up with the one at a time.  Many functions in hpgltools are quite chatty with
liberal usage of message(), as a result I will sm() this call to silence it.

```{r graph_metrics, fig.show = "hide"}
pasilla_metrics <- sm(graph_metrics(pasilla_expt, ma = TRUE, qq = TRUE))
summary(pasilla_metrics)
```

Now let us print some plots.

```{r print_graphs}
pasilla_metrics$libsize
## The library sizes range from 8-21 million reads, this might be a problem for
## some analyses, but it should be ok
pasilla_metrics$nonzero
## Ergo, the lower abundance libraries have more genes of counts == 0 (bottom
## left).
pasilla_metrics$boxplot
## And a boxplot downshifts them (but not that much because it decided to put
## the data on the log scale).
pasilla_metrics$density
## Similarly, one can see those samples are a bit lower with respect to density

## Unless the data is very well behaved, the rest of the plots are not likely to
## look good until the data is normalized, nonetheless, lets see
pasilla_metrics$corheat
pasilla_metrics$disheat
pasilla_metrics$pc_plot
## So the above 3 plots are pretty much the worst case scenario for this data.
```

# Normalize and replot

The most common normalization suggested by Najib is a cpm(quantile(filter(data))).
On top of that we often do log2() and/or a batch adjustment.
default_norm() does the first and may be supplemented with other arguments.

```{r normalize, fig.show = "hide"}
norm <- default_norm(pasilla_expt, transform = "log2")
norm_metrics <- graph_metrics(norm)
```

```{r show_norm}
norm_metrics$corheat
norm_metrics$smc
norm_metrics$disheat
norm_metrics$smd
## some samples look a little troublesome here.
norm_metrics$pc_plot
```

# Try a pairwise comparison

With the above metrics in mind, we may perform a pairwise comparison of the data.
By default, all_pairwise() performs every possible pairwise contrast, which in
the case is comprised of just treated vs. untreated.

```{r perform_pairwise, fig.show = "hide"}
pasilla_pairwise <- sm(all_pairwise(pasilla_expt))
pasilla_tables <- sm(combine_de_tables(
  pasilla_pairwise,
  excel = "pasilla_tables.xlsx"))
pasilla_sig <- sm(extract_significant_genes(
  pasilla_tables,
  excel = "pasilla_sig.xlsx"))
pasilla_ab <- sm(extract_abundant_genes(
  pasilla_pairwise,
  excel = "pasilla_abundant.xlsx"))
```

```{r de_pictures}
pasilla_tables[["plots"]][["untreated_vs_treated"]][["deseq_ma_plots"]]$plot
pasilla_tables[["plots"]][["untreated_vs_treated"]][["edger_ma_plots"]]$plot
pasilla_tables[["plots"]][["untreated_vs_treated"]][["limma_ma_plots"]]$plot
```

```{r goseq_test}
up_genes <- pasilla_sig[["deseq"]][["ups"]][["untreated_vs_treated"]]
down_genes <- pasilla_sig[["deseq"]][["downs"]][["untreated_vs_treated"]]
pasilla_go <- load_biomart_go(species = "dmelanogaster")$go
pasilla_length <- fData(pasilla_expt)[, c("ensembl_gene_id", "cds_length")]
colnames(pasilla_length) <- c("ID", "length")

pasilla_up_goseq <- simple_goseq(sig_genes = up_genes, go_db = pasilla_go, length_db = pasilla_length)
pasilla_up_goseq[["pvalue_plots"]][["bpp_plot_over"]]

pasilla_down_goseq <- simple_goseq(sig_genes = down_genes, go_db = pasilla_go,
                                   length_db = pasilla_length)
pasilla_down_goseq[["pvalue_plots"]][["bpp_plot_over"]]

pasilla_high_goseq <- simple_goseq(
    sig_genes = names(pasilla_ab[["abundances"]][["deseq"]][["high"]][["treated"]]),
    go_db = pasilla_go, length_db = pasilla_length)
pasilla_high_goseq[["pvalue_plots"]][["bpp_plot_over"]]

pasilla_low_goseq <- simple_goseq(
    sig_genes = names(pasilla_ab[["abundances"]][["deseq"]][["low"]][["treated"]]),
    go_db = pasilla_go, length_db = pasilla_length)
pasilla_low_goseq[["pvalue_plots"]][["bpp_plot_over"]]
```

```{r saveme}
pander::pander(sessionInfo())
message(paste0("This is hpgltools commit: ", get_git_commit()))
```
