---
title: "hpgltools examples using the fission dataset"
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
 html_document:
  code_download: true
  code_folding: show
  fig_caption: true
  fig_height: 7
  fig_width: 7
  highlight: default
  keep_md: false
  mode: selfcontained
  number_sections: true
  self_contained: true
  theme: readable
  toc: true
  toc_float:
    collapsed: false
    smooth_scroll: false
vignette: >
  %\VignetteIndexEntry{b-02_fission_data_exploration}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r options, include=FALSE}
## These are the options I tend to favor
library("hpgltools")
knitr::opts_knit$set(
    progress = TRUE,
    verbose = TRUE,
    width = 90,
    echo = TRUE)
knitr::opts_chunk$set(
    error = TRUE,
    fig.width = 8,
    fig.height = 8,
    dpi = 96)
options(
    digits = 4,
    stringsAsFactors = FALSE,
    knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size=10))
set.seed(1)
rmd_file <- "b-02_fission_data_exploration.Rmd"
```

```{r rendering, include=FALSE, eval=FALSE}
## This block is used to render a document from within it.
rmarkdown::render(rmd_file)

rmarkdown::render(rmd_file, output_format="pdf_document", output_options=c("skip_html"))

## Or to save/load large Rdata files.
hpgltools:::saveme()
hpgltools:::loadme()
rm(list=ls())
```

# Example hpgltool usage with a real data set (fission)

This document aims to provide further examples in how to use the hpgltools.

Note to self, the header has rmarkdown::pdf_document instead of html_document or html_vignette
because it gets some bullcrap error 'margins too large'...

## Setting up

Here are the commands I invoke to get ready to play with new data, including everything
required to install hpgltools, the software it uses, and the fission data.

```{r setup, include=TRUE}
## These first 4 lines are not needed once hpgltools is installed.
## source("http://bioconductor.org/biocLite.R")
## biocLite("devtools")
## library(devtools)
## install_github("elsayed-lab/hpgltools")
library(hpgltools)
require.auto("fission")
## I use the function 'sm()' to quiet loud functions.
tt <- sm(library(fission))
tt <- sm(data(fission))
```

## Data import

All the work I do in Dr. El-Sayed's lab makes some pretty hard
assumptions about how data is stored.  As a result, to use the fission
data set I will do a little bit of shenanigans to match it to the
expected format.  Now that I have played a little with fission, I
think its format is quite nice and am likely to have my experiment
class instead be a SummarizedExperiment.

```{r data_import}
## Extract the meta data from the fission dataset
meta <- as.data.frame(fission@colData)
## Make conditions and batches
meta$condition <- paste(meta$strain, meta$minute, sep=".")
meta$batch <- meta$replicate
meta$sample.id <- rownames(meta)
## Grab the count data
fission_data <- fission@assays$data$counts
## This will make an experiment superclass called 'expt' and it contains
## an ExpressionSet along with any arbitrary additional information one might want to include.
## Along the way it writes a Rdata file which is by default called 'expt.Rdata'
fission_expt <- create_expt(metadata=meta, count_dataframe=fission_data)
```

## Normalizing and exploring data

There are lots of toys we have learned to use to play with with raw
data and explore stuff like batch effects or non-canonical
distributions or skewed counts.  hpgltools provides some functionality
to make this process easier.  The graphs shown below and many more are
generated with the wrapper 'graph_metrics()' but that takes away the
chance to explain the graphs as I generate them.

```{r norm_explore}
## First make a bar plot of the library sizes in the experiment.
## Notice that the colors were auto-chosen by create_expt() and they should
## be maintained throughout this process
fis_libsize <- plot_libsize(fission_expt)
fis_libsize
## Here we see that the wild type replicate 3 sample for 15 minutes has fewer non-zero genes than all its friends.
fis_nonzero <- plot_nonzero(fission_expt, labels="boring", title="nonzero vs. cpm")
fis_nonzero
```

### An initial pca plot

In most cases, raw data does not cluster very well, lets see if that
is also true for the fission experiment. Assuming it doesn't, lets
normalize the data using the defaults (cpm, quantile, log2) and try
again.

```{r pca}
## Something in this is causing a build loop on travis...
## Unsurprisingly, the raw data doesn't cluster well at all...
fis_rawpca <- plot_pca(fission_expt, expt_labels=fission_expt$condition)
fis_rawpca$plot
## So, normalize the data
norm_expt <- sm(normalize_expt(fission_expt, transform="log2", norm="quant", convert="cpm"))
## And try the pca again
fis_normpca <- plot_pca(norm_expt, plot_labels="normal", title="normalized pca")
fis_normpca$plot

summary(fis_normpca)
```

In the final line of the preceeding block, I printed a summary of the return from plot_pca().
It contains the following information:

* pca:      The result from the fast.svd() call.
* plot:     The ggplot2 pca plot.
* table:    The metadata used to make the pca plot.
* res:      A table of the residual variance after each component by condition/batch.
* variance: A numeric list of the %variance remaining after each PC.

With that in mind, lets perform some more pca plots after normalizing the data and see how different
they look.

```{r normalized_pca}
normbatch_expt <- sm(normalize_expt(fission_expt, transform="log2", norm="quant", convert="cpm", batch="sva"))
fis_normbatchpca <- plot_pca(normbatch_expt, title="Normalized PCA with batch effect correction.")
fis_normbatchpca$plot
## ok, that caused the 0, 60, 15, and 30 minute samples to cluster nicely
## the 120 and 180 minute samples are still a bit tight

## pca_information provides some more information about the call to
## fast.svd that went into making the pca plot
fis_info <- pca_information(norm_expt, expt_factors=c("condition","batch","strain","minute"), num_components=6)
## The r^2 table shows that quite a lot of the variance in the data is explained by condition
head(fis_info$rsquared_table)
## We can look at the correlation between the principle components and the factors in the experiment
## in this case looking at condition/batch vs the first 4 components.
fis_info$pca_cor
## And p-values to lend some credence(or not to those assertions)
fis_info$anova_p

## Try again with batch removed data
batchnorm_expt <- sm(normalize_expt(fission_expt, batch="limma", norm="quant", transform="log2", convert="cpm"))
fis_batchnormpca <- plot_pca(batchnorm_expt, plot_title="limma corrected pca")
fis_batchnormpca$plot
test_pca <- pca_information(batchnorm_expt, expt_factors=c("condition","batch","strain","minute"), num_components=6)
```

Interesting, the batch normalized pca plot looks much the same as the
normalized. The variances are in fact pretty much the exact same...

## Look at the data distributions

We have some tools which provide visualizations of the distribution of
the data:

```{r distributions}
fission_boxplot <- sm(plot_boxplot(fission_expt))
fission_boxplot
sf_expt <- sm(normalize_expt(fission_expt, norm="sf"))
fission_boxplot <- sm(plot_boxplot(sf_expt))
fission_boxplot
tm_expt <- sm(normalize_expt(fission_expt, norm="tmm"))
fission_boxplot <- sm(plot_boxplot(tm_expt))
fission_boxplot
rle_expt <- sm(normalize_expt(fission_expt, norm="rle"))
fission_boxplot <- sm(plot_boxplot(rle_expt))
fission_boxplot
up_expt <- sm(normalize_expt(fission_expt, norm="upperquartile"))
fission_boxplot <- sm(plot_boxplot(up_expt))
fission_boxplot

fission_density <- plot_density(norm_expt)
fission_density
fission_density <- plot_density(sf_expt)
fission_density
fission_density <- plot_density(tm_expt)
fission_density

compare_12 <- plot_qq_plot(fission_expt, x=1, y=2)
compare_12$log
```

## See how they cluster

Ok, so we can further check out how the data cluster with respect to
one another...

```{r clustering}
fission_cor <- plot_corheat(norm_expt)
fission_cor$plot
fission_cor <- plot_corheat(batchnorm_expt)
fission_cor$plot
fission_dis <- plot_disheat(norm_expt)
fission_dis$plot
fission_dis <- plot_disheat(batchnorm_expt)
fission_dis$plot
```

# variancePartition

A new toy I have been introduced to is variancePartition, let us try that with this data.

```{r variancePartition}
test_varpart <- varpart(fission_expt, predictor=NULL, factors=c("condition","batch"))
test_varpart$percent_plot
test_varpart$partition_plot

## Here, let us test the variance contributed by strain, time, and replicate.
test_varpart <- varpart(fission_expt, predictor=NULL, factors=c("condition", "strain", "minute", "replicate"))
test_varpart$percent_plot
test_varpart$partition_plot
```

YAY!

[index.html](index.html)

```{r sysinfo, results='asis'}
pander::pander(sessionInfo())
```
